### 2016-11-24 Manuscript/SW_early_life_telomeres.Rmd
Eds

**Spatiotemporal variation in lifelong telomere dynamics in the Seychelles warbler**

@@

2. The Seychelles warbler (*Acrocephalus sechellensis*) is an excellent model system for disentangling the causes and consequences of individual variation in telomere dynamics. Here we study spatiotemporal variation in lifelong telomere dynamics in the Seychelles warbler. We combine long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from multiple cohorts spanning `r max(dd$CatchYear)-min(dd$CatchYear)` years. We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment.

@@

#####Statistical analyses
We performed all statistical analyses using R. We first explored the relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014]. Following a similar approach to Fairlie *et al.* [-@Fairlie2016], we compared a selection of models fitting different relationships between RTL and age. We created models where age was fitted as a linear covariate, where age was log-transformed, and where age was fitted as a factor. We also created models including birth year (cohort) as a factor, and ones which included interactions between cohort and each measure of age. All fitted models are included in Table 1. As random effects we included individual ID, catch year and qPCR plate ID. Models were compared using AIC.

Using the longitudinal data, we tested whether telomores shorten with age in individuals, using a LMM of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009]. Because we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error. Using all birds (adults and juveniles) with at least two telomere measurement, we calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample). We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (i.e. completely seperate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples). We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age. We first compared the variance in $\Delta$RTL within and among individuals using a Levene's test. Then, to separately test whether the extent of telomere shortening and lenghtening within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which shortening ($\Delta$RTL < 0) and lengthening $\Delta$RTL ($\Delta$RTL > 0) occurred, and tested whether $\Delta$RTL among samples was significantly different between within-sample and across-sample measurements, using Wilcoxon tests.

We used LMMs to test how variation in early-life environmental and social conditions influenced telomere length within cohorts. We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, annual food availability, territory quality, population density (an annual measure estimated from the summer breeding census) and the number of birds and the number of helping  subordinate birds present in the territory. As random effects we included catch year and individual identity. We report model estimates and confidence intervals for all effects included in the full model. To obtain an estimate of explanatory power, we calculated marginal R^2 (just incoporatng fixed effects) and conditional R^2 (incorporating fixed and random effects) following Nakagawa & Schielzeth [-@Nakagawa2013] and Johnson [-@Johnson2014], respectively As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012]. Model selection was performed using the full model described above and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL as the response variable, and excluding the plate ID random effect (as each measurement of $\Delta$RTL was based on two or more measurements, and so run on multiple plates).

###Results
We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers. Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively. Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`-`r round2(rptall$CI_emp[2])`)

#####Telomere dynamics and age
Model comparison

Within individuals, longitudinal data showed that RTL measurements taken from different timepoints were weakly, but significantly, positvely correlated (R^2^ = `r round3(summary(lm(RTL1~RTL,data = dd3))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL,data = dd3))$coefficients[2,4])`; Fig. 3A). Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our $\Delta$RTL measurments were positive, and thus indicative of telomere lengthening. To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 3B). Importantly, splitting the longitudinal data into instances of shortening ($\Delta$RTL < 0) and lenghtning ($\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: W = `r round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$statistic,0)`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (W = `r as.character(round(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$statistic,0))`; P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 3B).

#####Spatiotemporal variation in telomere dynamics and the environment
A LMM revealed that age, tarsus length, sex and annual variation in food availability were significantly related to RTL (Table 1). Tarsus length was negatively related to RTL and males had longer telomeres than females (Table 1; Fig. 4A), while insect availability was positively related to RTL (Table 1; Fig. 4B). The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`) The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig. S1). One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1). In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL increased with tarsus length in males, but decreased in females (Fig. 4A). No social or ecological environmental variables were significant predictors of $\Delta$RTL, regardless of whether we interpreted the full model  (Table 2), or used model averaging (Table S2; Fig. S2).

###Discussion
Here we use a long-term, multi-cohort dataset to assess the relationships between the spatiotemporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers. We found that telomeres shorten over time, and that telomere loss is greatest in early life. However, bouts of apparent telomere lengthening occur, and the extent of this lengthening cannot be explained solely by qPCR measurement error. We also found that telomere dynamics vary among sexes and over time, with temporal variation in telomere dynamics related to fluctuations in food availability.

Our study adds the the now substantial body of literature from humans and wild animals, showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2016]. However, using longitudinal data we found that within-individual telomere measurments were only weakly correlated, and that consistent telomere shortening only occurs early in life, after which both shortening and lengthening occurs. This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which juvenile and adult telomere length were highly correlated, and individual-level telomere shortening clearly occurs throughout the juvenile period, and into adulthood. However, in accordance with our study, a recent study on a wild population of Soay sheep (*Ovis aries*) found only a very weak correlation between juvenile and adult RTL, and a highly complex pattern of telomere dynamics across the lifespan [@Fairlie2016].

@@

Perhaps the clearest result from our study is that RTL, among cohorts, and among sampling years. A few studies have now shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016]. The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods. Moreover, because of the large number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in food availability (Fig. 4B). If cohort-level variation in telomere dynamics continues to be found in other systems, this has potential consequences for our understanding of telomere dynamics in natural populations. In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process. More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.

Despite substantial cohort-level variation in RTL, within cohorts we found no evidence that any of the social or environmental variables we measured affect juvenile RTL. This is surprising, as a poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival [@VandeCrommenacker2011a; @Brouwer2012]. An explanation for this could be that temporal variation in RTL in our data masks our ability to detect spatial trends. However, this is unlikely as we controlled for cohort ID, and within-cohort variation in RTL greatly exceeds among-cohort variation (Fig. 1C). A second possibility is that effects are generally weak and levels of noise in our telomere measurements and/or ecological data preclude detection of significant effects. While noise in telomere measurements and problems with power are a problem in any study using qPCR to measure telomere length [@Nussey2014], our sample sizes and levels of repeatability compare favourably with other studies of natural populations that have found significant effects of the early-life environment on telomere length [e.g. @Fairlie2016]. More likely is that early-life RTL in the Seychelles warbler is explained by a set of variables that we have not measured. A key question to be addressed is the extent to which early-life RTL reflects inheritance [e.g. @Becker2015; @Asghar2014]. This is important, as inheritance may capture a large amount of unexplained variation in early-life RTL within our dataset. Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence.

While the relationship between mortality and telomeres in adults has been established for some time [@Cawthon2003], only recently has the link between later-life survival and early-life telomere dynamics been studied. In captive zebra finches, juvenile telomere length predicts late-life survival [@Heidinger2012], and in wild bird populations survival to the nestling phase [@Watson2015], and survival to adulthood [@Boonekamp2014], have been linked with early-life telomere dynamics. It is worth noting that here we have considered survival, but telomere length and shortening in early life may also be linked to other components of fitness [reviewed in @Monaghan2014]. Reproductive senescence occurs in the Seychelles warbler [reviewed in @Hammers2015], making this system well suited to examining how telomere length predicts lifetime reproductive success.  We expect that by gaining a fuller understanding of telomere dynamics in natural populations, the fields of life-history evolution and evolutionary ecology will be greatly enhanced.

>>>>

**Spatiotemporal variation in lifelong telomere dynamics in a long-term ecological study**

@@

2. The Seychelles warbler (*Acrocephalus sechellensis*) is an excellent model system for disentangling the causes and consequences of individual variation in telomere dynamics. Here we study spatiotemporal variation in lifelong telomere dynamics in the Seychelles warbler. We combine long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatiotemporal variation in the social and ecological environment.

@@

#####Statistical analyses
We performed all statistical analyses using R. We first explored the relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014]. Following a similar approach to Fairlie *et al.* [-@Fairlie2016], we compared a selection of models fitting different relationships between RTL and age. We created models where age was fitted as a linear covariate, where age was log-transformed, and where age was fitted as a factor. We also created models including birth year (cohort) as a factor, and ones which included interactions between cohort and each measure of age. All fitted models are included in Table 1. As random effects we included individual ID, catch year and qPCR plate ID. Models were compared using AIC with correction for finite sample size (AICc). Using the longitudinal data, we tested whether telomores shorten with age in individuals, using LMMs of RTL and $\Delta$age [a longitudinal measure based on within-subject centring; @VandePol2009]. We calculated $\Delta$age using log and polynomial transformed age data, and carried out model selection as above, with the exception that we did not model $\Delta$age as a factor (due to a lack of discreet groupings).

Because, when examining the distribution of $\Delta$RTL, we observed apparent telomere lengthening in our data, we used the entire Seychelles warbler dataset to test whether this lengthening could be explained by measurement error. Using all birds (adults and juveniles) with at least two telomere measurement, we calculated $\Delta$RTL for each pair of samples (for example, for a bird with three longitudinal measurements throughout their life we calculated the change in RTL between the first and second sample and between the second and third sample). We then compared these longitudinal changes in telomere length to sampling error, using all samples where we had at least two repeat telomere measurements for the same sample (i.e. completely seperate reactions run on separate plates: N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples). We calculated $\Delta$RTL between pairs of repeat measurements within samples in exactly the same way as for across samples, except that repeat measurements were ordered by the date at the qPCR was run rather than by age. We first compared the variance in $\Delta$RTL within and among individuals using a Levene's test. Then, to separately test whether the extent of telomere shortening and lenghtening within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which shortening ($\Delta$RTL < 0) and lengthening $\Delta$RTL ($\Delta$RTL > 0) occurred, and tested whether $\Delta$RTL among samples was significantly different between within-sample and across-sample measurements, using Wilcoxon tests.

We used LMMs to test how variation in early-life environmental and social conditions influenced telomere length within cohorts. We created a full model with RTL as a response variable, alongside the following explanatory variables: age, tarsus length, body mass, sex, annual food availability, territory quality, population density (an annual measure estimated from the summer breeding census) and the number of birds and the number of helping  subordinate birds present in the territory. Random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and allowed the slope of the age effect to vary among cohorts. We report model estimates and confidence intervals for all effects included in the full model. To obtain an estimate of explanatory power, we calculated marginal R^2 (just incoporatng fixed effects) and conditional R^2 (incorporating fixed and random effects) following Nakagawa & Schielzeth [-@Nakagawa2013] and Johnson [-@Johnson2014], respectively As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012]. Model selection was performed using the full model described above and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(dd3)`) we repeated the above analyses of telomere dynamics, replacing telomere length with $\Delta$RTL as the response variable, and excluding the random effects for plate ID (as each measurement of $\Delta$RTL was based on two or more measurements, and so run on multiple plates) and cohort ID (as longitudinal telomere dynamics did not differ among cohorts; see results).

###Results
We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively. Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`-`r round2(rptall$CI_emp[2])`).

#####Telomere dynamics and age amnong cohorts
We first tested how RTL was related to age among cohorts using a model selection approach. We found that the top model, by some distance, contained a log age, cohort ID, and a log age x cohort interaction (Table 1A). The second best model contained log age and cohort ID, with no interaction term, although this was substantially poorer in terms of model fit ($\Delta$AICc > 10), and all other models were much poorer still ($\Delta$AICc > 35; Table 1). The log-linear relationship between RTL and age couls be clearly seen in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring early in life (Fig. 1A). There was also substantial variation in RTL among cohorts, with no obvious trend over time (Fig. 1B). There was a negative relationship between RTL and age in almost all `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between log age and RTL varied substantially among cohorts (Fig.1C).

A within-individual analysis of RTL and age revealed that the top model explaining RTL contained $\Delta$log age, which reflects within-individual changes in log-transformed age (Table 1B). Models including cohort ID and cohort x age interactions were substantially poorer than models only containing age (Table 1B). RTL dereased with $\Delta$log age (estimate = `r round3(fixef(RTL_DeltaLogAge)['DeltaLogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_DeltaLogAge)['DeltaLogAge',],lessthan = F),collapse = ', ')`), confirming that within-individual telomere shortening occurs across the Seychelles warbler dataset.

There was a hint of a positive correlation between telomere measurements taken from different timepoints (Fig. 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not quite significant (estimate = `r round3(fixef(RTL_RTL1)['RTL1'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`). Although both or cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increaseed in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of cases (Fig. 2A), which indicative of telomere lengthening. To test whether the apparent telomere lengthening in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual (see methods for details). We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 2B). Importantly, splitting the longitudinal data into instances of shortening (i.e. $\Delta$RTL < 0) and lenghtning (i.e. $\Delta$RTL > 0) revealed that not only did we observe significantly more shortening within individuals compared to within samples (Wilcoxon test: P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also significantly more lengthening (P  `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 2B).

#####Telomere dynamics and the environment
A LMM revealed that age, tarsus length, sex and insect abundance were significantly related to RTL (Fig. 3A). Tarsus length was negatively related to RTL and males had longer telomeres than females (Fig. 3B), while insect abundance was positively related to RTL (Fig. 3C). The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`) The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig. S1). One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1). In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL increased with tarsus length in males, but decreased in females (Fig. 4A). No social or ecological environmental variables were significant predictors of $\Delta$RTL using the full model  approach (Table S2). Using model selection, we found that the top model explaining $\Delta$RTL contained age and population density (Table S3). $\Delta$RTL was positively related to age, consistent with telomere shortening being highest in early life, and negatively related to population density; however, in both instances model averaged confidence intervals overlapped zero (Fig. S2).

###Discussion
Here we use a long-term, multi-cohort dataset to assess the relationships between the spatiotemporal variation in the ecological environment and lifelong telomere dynamics in a closed population of Seychelles warblers. We found that telomere length decreases with age, and that this decrease is greatest in early life. Telomere length decreased with age in almost all `r length(table(dd$Cohort)` cohorts studied, but telomere length, and the rate of decrease of telomere length with age, varied substantially among cohorts. Despite an overall trend for telomere shortening in the Seychelles warbler, we found that bouts of apparent telomere lengthening occur within individuals, and the extent of this lengthening cannot be explained solely by qPCR measurement error. Finally, we found that telomeres are related to tarsus length in a sex-specific manner, and that telomer length is positively associated with temporal fluctuations in food availability.

Our study adds the the now substantial body of literature from humans and wild animals showing that telomere length decreases with age, and that this decrease is most rapid in early life [e.g. @Frenck1998; @Haussmann2003; @Heidinger2012; @Fairlie2016]. However, we found that telomere shortening was only consistently observed early in life, after which both shortening and lengthening occurred. Importantly, these increases were observed in longitudinal as well as cross-sectional data, suggesting that selective disappearance does not explain this pattern entirely. Longitudinal increases in measured telomere length have been observed in humans and [] wild animals [], but whether these observations reflect true elongation or measurement error has been debated.

We are aware of two previously described methods that statistically identify whether telomere elongation occurs within a population. One approach uses individuals with at least three measurements to differentiate between hypothesis of consistent lenghtening and a combination of shortening and lengthening [@Simons2014]. Another study found that the likelihood of detectening increases in telomere length was related to the follow-up time between sampling events, which they argued is inconsistent with telomere elongation [@Steenstrup2013]. However both of these approaches assume the telomere elongation within individuals are consistent over time. We suggest that telomere elongation could occur at specific, short periods throughout individual lifespans, against a backdrop of overall telomere shortening.

Consistent with a dynamic pattern of telomere shortening with age, we found that within-individual telomere measurments were only weakly correlated. This is in contrast to other avian studies where birds were reared in laboratory [@Heidinger2012] or nest-box [@Boonekamp2014] conditions, in which within-individual telomere length measurements were highly correlated, and individual-level telomere shortening  occured throughout the juvenile period and into adulthood. However, a recent study on a wild population of Soay sheep (*Ovis aries*) found only a very weak correlation between within-individual telomere measurements [@Fairlie2016], and we anticipate that a similar pattern may be found in many other wild populations.

Perhaps the clearest result from our study is that RTL varied among cohorts. A few studies have now shown that temporal variation in telomere dynamics occurs in natural populations, although these have been limited in the number of seasons [@Mizutani2013; @Watson2015; @Fairlie2016]. The long-term Seychelles warbler dataset has allowed us to show that temporal variation in telomere dynamics can occur over substantial time periods. Moreover, because of the large number of cohorts we have been able to show statistically that that temporal variation in RTL is related to variation in food availability. If cohort-level variation in telomere dynamics continues to be found in other systems, this has potential consequences for our understanding of telomere dynamics in natural populations. In particular, our findings suggest that the telomere dynamics of a population at a given point in time represent a snapshot of a temporally varying process. More research is now needed within and across multiple cohorts and populations to better understand how how and why population-level telomere dynamics vary over space and time.

@@

Despite finding associations between the environment and telomere dynamics, we should bear in mind that our social and ecological variables explained a small proportion of the variance in RTL. A poor social and ecological environment is known to be detrimental to juvenile Seychelles warblers, both in terms of oxidative stress and later life survival [@VandeCrommenacker2011a; @Brouwer2012]. An explanation for this could be that temporal variation in RTL in our data masks our ability to detect spatial trends. However, this is unlikely as we controlled for cohort ID, and within-cohort variation in RTL greatly exceeds among-cohort variation (Fig. 1C). A second possibility is that effects are generally weak and levels of noise in our telomere measurements and/or ecological data preclude detection of significant effects. While noise in telomere measurements and problems with power are a problem in any study using qPCR to measure telomere length [@Nussey2014], our sample sizes and levels of repeatability compare favourably with other studies of natural populations that have found significant effects of the early-life environment on telomere length [e.g. @Fairlie2016]. More likely is that early-life RTL in the Seychelles warbler is explained by a set of variables that we have not measured. A key question to be addressed is the extent to which early-life RTL reflects inheritance [e.g. @Becker2015; @Asghar2014]. This is important, as inheritance may capture a large amount of unexplained variation in early-life RTL within our dataset. Future research on the Seychelles warbler will examine, in a quantitative genetic framework, how genetic and environmental components, and their interactions, affect telomere dynamics and senescence.

### 2016-11-24 Manuscript/SW_early_life_telomeres_SOM.Rmd
EDS
Eds

###Supplementary information for Spurgin *et al.* "Telomeres reveal silver spoon effects in a wild population"

@@

#Model averaging

@@

**Table S2** Model selection table for analysis of factors affecting $\Delta$RTL in Seychelles warbler nestlings, using longitudinal data. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here. TQ = territory quality.

```{r Table S2}

@@

@@

tableS2 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
colnames(tableS2) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS2) <- NULL
tableS2$Model[tableS2$Model == ''] <- 'Null model'
tableS2[,3:5] <- lapply(tableS2[,3:5],as.numeric)
kable(tableS2,digits=2)

@@

#Sort out row/col names
nams <- rownames(table2)
nams[nams == 'z.GroupSize'] <- 'Group Size'
nams[nams == 'c.Sex'] <- 'Sex (males)'
nams[nams == 'c.Helper'] <- 'Helpers'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.LogAge'] <- 'Log age'
nams[nams == 'z.Tarsus'] <- 'Tarsus'
nams[nams == 'z.Insect'] <- 'Insect abundance'
nams[nams == 'z.CatchMonth'] <- 'Month'
nams[nams == 'z.Density'] <- 'Density'
nams[nams == 'z.TQ'] <- 'Territory quality'

>>>>

###Supplementary information for Spurgin *et al.* "Spatiotemporal variation in lifelong telomere dynamics in a long-term ecological study"

@@

@@

**Table S2** Full model of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data.
```{r Table S2}

#Run the model
CIs <- CIwald(DeltaRTL_Full)
t1b <- data.frame(Variable = row.names(CIs)[3:nrow(CIs)],
                  Estimate = fixef(DeltaRTL_Full),
                  LCI = CIs[3:nrow(CIs),1],
                  UCI = CIs[3:nrow(CIs),2],
                  row.names = NULL,
                  stringsAsFactors = F)

nams <- t1b$Variable
nams[nams == 'z.LogAge'] <- 'Age'
nams[nams == 'z.Tarsus'] <- 'Tarsus length'
nams[nams == 'c.Helper'] <- 'Number helpers'
nams[nams == 'z.GroupSize'] <- 'Group size'
nams[nams == 'c.Sex'] <- 'Sex (males)'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.Insect'] <- 'Insect abundance'
nams[nams == 'z.TQ'] <- 'Territory quality'
nams[nams == 'z.Density'] <- 'Density'

t1b$Variable <- nams

kable(t1b,digits = 3)

**Table S3** Model selection table for analysis of factors affecting $\Delta$RTL in Seychelles warblers, using longitudinal data. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here. TQ = territory quality.

```{r Table S3}

@@

nams[nams == 'z.GroupSize'] <- 'Group size'
nams[nams == 'c.Sex'] <- 'Sex (males)'
nams[nams == 'c.Helper'] <- 'Helpers'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.LogAge'] <- 'Log age'
nams[nams == 'z.Tarsus'] <- 'Tarsus'
nams[nams == 'z.Insect'] <- 'Insect abundance'
nams[nams == 'z.CatchMonth'] <- 'Month'
nams[nams == 'z.Density'] <- 'Density'
nams[nams == 'z.TQ'] <- 'Territory quality'

@@

tableS3 <- data.frame(Model = apply(t1,1,getterms,c(2:10)),t1[,c('df','AICc','delta','weight')],stringsAsFactors = FALSE)
colnames(tableS3) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS3) <- NULL
tableS3$Model[tableS3$Model == ''] <- 'Null model'
tableS3[,3:5] <- lapply(tableS3[,3:5],as.numeric)
kable(tableS3,digits=2)

@@

#Sort out row/col names
nams <- rownames(table2)
nams[nams == 'z.GroupSize'] <- 'Group size'
nams[nams == 'c.Sex'] <- 'Sex (males)'
nams[nams == 'c.Helper'] <- 'Helpers'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.LogAge'] <- 'Log age'
nams[nams == 'z.Tarsus'] <- 'Tarsus'
nams[nams == 'z.Insect'] <- 'Insect abundance'
nams[nams == 'z.CatchMonth'] <- 'Month'
nams[nams == 'z.Density'] <- 'Density'
nams[nams == 'z.TQ'] <- 'Territory quality'
